An exposition on the propriety of restricted Boltzmann machines

### 2016-09-03 writing/draft.Rmd
add Vardeman's edits for abstract and intro

---
title: "An exposition on the propriety of restricted Boltzmann machines"
author: "Andee Kaplan, Daniel Nordman, Stephen Vardeman"
date: "August, 2016"
output: 
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    number_sections: yes
bibliography: references/refs.bib
abstract: |
  A restricted Boltzmann machine (RBM) is an undirected graphical model constructed for discrete or continuous random variables, specified as having two layers, a hidden and a visible layer with no conditional dependency within a layer. In recent years, RBMs have risen to prominence due to their connection to deep learning - by treating a hidden layer of one RBM as the visible layer in a second RBM a deep architecture can be created. The method is claimed to have the ability to learn very complex and rich structures in data, making these models attractive for supervised learning. However, the generative behavior of RBMs has largely been unexplored. In this paper, we discuss the relationship between parameter specification and the prevalence of degenerate models as well as an exposition of the difficulties in fitting such highly flexible models in the binary case of RBMs.
---

@@

The data mining and machine learning communities have recently found a great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBM). (**TODO: deep learning, deep RBM references.**) A RBM is an undirected graphical model constructed for discrete or continuous random variables, specified as having two layers, a hidden and a visible layer with no conditional dependency within a layer. These models are typically used with success in classification of images. (**TODO: classification references.**) However, the model properties are largely unexplored in the literature and the fitting methodology remains heuristic based and abstruse. In this paper, we attempt to give a thorough understanding of the model and its behavior from the context of statistical model theory as well as explore the possibility of a rigorous fitting methodology. Through our study, we find the RBM model to be deficient in two fundamental ways.

First, the models are unsatisfactory as a conceptualization of how data is generated. According to Fisher, the aim of a statistical model is to represent data in a compact way [@fisher1922mathematical]. Neyman and Box agree with this assessment and go further to state that a model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model, @box1967discrimination]. In the case of RBMs, these models often fail to generate data that resembles realistic data. Thus, providing an unsatisfactory conceptualization of the data generation process. Additionally, it is evidenced that RBMs easily exhibit instability in the parameter space. In practice, this is seen when a single pixel change in an image results in a wildly different classification. These model impropriety measures has been documented in other deep architectures recently as well [@szegedy2013intriguing, @nguyen2014deep]. We investigate these phenomena in RBMs in section \ref{explorations} through simulations of small, manageable examples.

Secondly, the fitting of these models is problematic. As the size of these models grows, a computationally intractable situation is created quickly when fitting with either maximum likelihood or through Bayesian methods. However, even in models small enough to feasibly fit, there are unique challenges due to the overparameterization of RBMs. In section \ref{bayesian-model-fitting} we try three Bayesian modelling techniques, each with the goal of avoiding parts of the parameter space that yield improper models. As could be expected, with computational complexity comes an increase in accuracy, but at the cost of feasibility. Additionally, because these models are overparametrized, any honest fitting method will reproduce the empirical distribution. Thus, there is no smoothing achieved with a RBM that is fit using a rigorous method. We are therefore very skeptical that any model built using these structures, like a deep Boltzmann machine, can achieve prediction or inference in a principled way.

This paper is structured as follows, section \ref{background} gives a definition of the RBM including the joint distribution and shows the connection to deep learning. Additionally, measures of model impropriety and methods of detecting them are defined. Section \ref{explorations} details our explorations into the model behavior and propriety (or lack thereof) of the RBM. We discuss three Bayesian fitting techniques with the goal of avoiding model impropriety in section \ref{bayesian-model-fitting} and wrap up with a discussion in section \ref{discussion}. Additionally, there are two accompanying appendices with auxiliary proofs relating to the choice of data encoding and parameter behavior during fitting with incomplete training data in section \ref{appendix}.

@@

### Connection to Deep Learning
RBMs have risen to prominence in recent years due to their connection to deep learning [see @hinton2006fast, @salakhutdinov2012efficient, or @srivastava2013modeling for examples]. By stacking multiple layers of RBMs in a deep architecture, the method's proponents claim the ability to learn "internal representations that become increasingly complex, which is considered to be a promising way of solving object and speech recognition problems" [@salakhutdinov2009deep, pp. 450]. The stacking is achieved by treating a hidden layer of one RBM as the visible layer in a second RBM, and so on until the desired architecture is created.

>>>>

---
title: "An exposition on the propriety of restricted Boltzmann machines"
author: "Andee Kaplan, Daniel Nordman, Stephen Vardeman"
date: "August, 2016"
linestretch: 2 
output: 
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    number_sections: yes
bibliography: references/refs.bib
abstract: |
  A restricted Boltzmann machine (RBM) is an undirected graphical model constructed for discrete or continuous random variables, with two layers, one hidden and one visible, with no conditional dependency within a layer. In recent years, RBMs have risen to prominence due to their connection to deep learning. By treating a hidden layer of one RBM as the visible layer in a second RBM a deep architecture can be created. RBMs are thought to have the ability to encode very complex and rich structures in data, making them attractive for supervised learning. However, the generative behavior of RBMs has largely been unexplored. In this paper, we discuss the relationship between parameter RBM specification and the prevalence of degenerate models as well as the difficulties that arise in fitting such highly flexible models in the binary case.
---

@@

The data mining and machine learning communities have recently found a great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBMs). (**TODO: deep learning, deep RBM references.**) A RBM is an undirected probabilistic graphical model (for discrete or continuous random variables) with two layers, one hidden and one visible, with no conditional dependency within a layer (**TODO: RBM references.**). These models have reportedly been used with success in classification of images. (**TODO: classification references.**) However, the model properties are largely unexplored in the literature and the commonly cited fitting methodology remains heuristic-based and abstruse. In this paper, we attempt to provide steps toward a thorough understanding of the model and its behavior from the perspective of statistical model theoryand then explore the possibility of a rigorous fitting methodology. We find the RBM model class to be deficient in two fundamental ways.

First, the models are unsatisfactory as a conceptualization of how data is generated. According to Fisher, the aim of a statistical model is to represent data in a compact way [@fisher1922mathematical]. Neyman and Box agree with this assessment and go further to state that a model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model, @box1967discrimination]. RBMs often fail to generate data that resemble realistic data and thus an unsatisfactory conceptualization of the data generation process. Additionally, we find that RBMs easily exhibit a kind of instability in the parameter space. In practice, this is seen when a single pixel change in an image results in a wildly different classification. Such model impropriety issues have  been documented in other deep architectures recently [@szegedy2013intriguing, @nguyen2014deep]. We investigate these phenomena in RBMs in section \ref{explorations} through simulations of small, manageable examples.

Secondly, the fitting of these models is problematic. As the size of these models grows, both maximum likelihood and Bayesian methods of fitting quickly being intractable. But even when fitting is feasible, there are unique challenges due to the overparameterization of RBMs. In section \ref{bayesian-model-fitting} we compare three Bayesian modelling techniques, each proposed with the goal of avoiding parts of the RBM parameter space that yield improper models. As could be expected, with computational complexity comes an increase in fitting accuracy, but at the cost of feasibility. Additionally, because these models are highly overparametrized, any fully principled fitting method will reproduce the empirical distribution of most realistic training sets. There is no "smoothing" achieved with a RBM fit using a rigorous method. We are therefore led to be skeptical that any model built using these structures, like a deep Boltzmann machine, can achieve prediction or inference in a principled way.

This paper is structured as follows, section \ref{background} gives a definition of the RBM including the joint distribution of hidden and visible variables and shows the connection to deep learning. Additionally, measures of model impropriety and methods of detecting them are defined. Section \ref{explorations} details our explorations into the model behavior and propriety (or lack thereof) of the RBM class. We discuss three Bayesian fitting techniques with the goal of avoiding model impropriety in section \ref{bayesian-model-fitting} and wrap up with a discussion in section \ref{discussion}. Additionally, there are two accompanying appendices with auxiliary proofs relating to the choice of data encoding and parameter behavior during fitting with incomplete training data in section \ref{appendix}.

@@

### Connection to Deep Learning

RBMs have risen to prominence in recent years due to their connection to deep learning [see @hinton2006fast, @salakhutdinov2012efficient, or @srivastava2013modeling for examples]. By stacking multiple layers of RBMs in a deep architecture, the method's proponents claim the ability to learn "internal representations that become increasingly complex, which is considered to be a promising way of solving object and speech recognition problems" [@salakhutdinov2009deep, pp. 450]. The stacking is achieved by treating a hidden layer of one RBM as the visible layer in a second RBM, and so on until the desired architecture is created.
